```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

# Week 2 - Exponential population growth {-} 

In part 1, we will solve the differential equation for exponential population growth and visualize how the population sizes change over time. 

<br>

**Part 1 - Numerical solution using the package "deSolve"**

Two main phases:

(1) Model specification: specify the structure of differential equation model

(2) Model application: set the time steps, initial population size and model parameters (e.g., intrinsic population growth rate *r*), and then solve the equation model

<style>

p span#aaa:before { 
  content: "(1) "; 
  display: inline-block;
  margin-left: -1.5em;
  margin-right: 0.3em;
}

p span#bbb:before { 
  content: "(2) "; 
  display: inline-block;
  margin-left: -1.5em;
  margin-right: 0.3em;
}

d-article table.lightable-paper {
  margin-bottom: 0px; 
}

</style>

Consider the model
\[
\frac{dN}{dt} = rN
\]
where $N$ is population size and $r$ is the intrinsic growth rate.
```{r}
###### part 1 ######
# install.packages("deSolve")
library(deSolve)

### (1) Model specification
exponential_model <- function(times, state, parms) {
  with(as.list(c(state, parms)), {
    dN_dt = r*N  # exponential growth equation
    return(list(c(dN_dt)))  # return the results
  })
}
```

Set the time steps, initial population size and model parameters.
```{r}
### (2) Model application
times <- seq(0, 10, by = 0.1)  # time steps to integrate over
state <- c(N = 10)  # initial population size
parms <- c(r = 1.5)  # intrinsic growth rate
```

Solve the equation by `ode()` numerically.
```{r}
# run the ode solver
pop_size <- ode(func = exponential_model, times = times, y = state, parms = parms)

# take a look at the results
head(pop_size)

```

Visualization
```{r}
plot(N ~ time, data = pop_size)
plot(N ~ time, data = pop_size, log = "y")
```


<br>

**Part 2 - Comparing different ode solvers**
<br>
In default of `ode()`, the equations are solved by LSODA method. We can change the method by modifying the argument `method` in `ode()`. 

```{r}

###### part 2 ######

times <- seq(0, 10, by = 0.1)  # time steps to integrate over
state <- c(N = 10)  # initial population size
parms <- c(r = 1.5)  # intrinsic growth rate
# default: LSODA
pop_size <- ode(func = exponential_model, times = times, y = state, parms = parms)

# Euler's method
pop_size_1 <- ode(func = exponential_model, times = times, y = state, parms = parms, method = "euler")

# compare different method
par(mfrow = c(1,2))
plot(N ~ time, data = pop_size, main = "LSODA")
curve(state[1]*exp(parms[1]*x), times[1], times[length(times)], col = "red", add = T) # correct curve
plot(N ~ time, data = pop_size_1, main = "Euler")
curve(state[1]*exp(parms[1]*x), times[1], times[length(times)], col = "red", add = T) # correct curve

# compare different method in log transformation
par(mfrow = c(1,2))
plot(N ~ time, data = pop_size, log = "y", main = "LSODA")
curve(state[1]*exp(parms[1]*x), times[1], times[length(times)], col = "red", add = T) # correct curve
plot(N ~ time, data = pop_size_1, log = "y", main = "Euler")
curve(state[1]*exp(parms[1]*x), times[1], times[length(times)], col = "red", add = T) # correct curve
```

<br>

**Part 3 - Solving exponential growth model with fluctuating growth rate**
<br>
Consider the model
\[
\frac{dN}{dt} = r(t)N \ \text{, } r(t) = \overline{r} + \frac{\sigma}{2}\sin(\omega t+\phi)
\]
where $\overline{r}$, $\omega$ and $\phi$ are constants.
The analytical solution of the ode model is 
\[
N(t) = N_0\exp\{\overline{r}t - \frac{\sigma}{2}[\cos(\omega t + \phi) - \cos(\phi)]\}
\]

```{r}
###### part 3 ######
### Model specification
exponential_model_fluc <- function(times, state, parms) {
  with(as.list(c(state, parms)), {
    dN_dt = (r_hat + sigma*sin(omega*times + phi)/2)*N  # exponential growth equation
    return(list(c(dN_dt)))  # return the results
  })
}
```


```{r}
### Parameters
times <- seq(0, 10, by = 0.1)  # time steps to integrate over
state <- c(N = 10)  # initial population size
parms <- c(r_hat = 1.5, sigma = 10, omega = 2*pi, phi = 0)  # intrinsic growth rate
```
Plot $r(t)$
```{r}
### Fluctuating growth rate
r = parms[1] + parms[2]/2*sin(parms[3]*times + parms[4])
plot(r ~ times, type = "l")
```

```{r}
### Solving model
pop_size <- ode(func = exponential_model_fluc, times = times, y = state, parms = parms)

### Plotting
plot(N ~ times, data = pop_size)
curve(state[1]*exp(parms[1]*x - parms[2]/2/parms[3]*(cos(parms[3]*x + parms[4]) - cos(parms[4]))), add = T) # correct curve
plot(N ~ times, data = pop_size, log = "y")
curve(state[1]*exp(parms[1]*x - parms[2]/2/parms[3]*(cos(parms[3]*x + parms[4]) - cos(parms[4]))), add = T) # correct curve
```
